\documentclass[11pt]{book}
\usepackage{fancyhdr}
\usepackage{multicol}
\usepackage{makeidx}
\usepackage[hidelinks]{hyperref}
\usepackage{refcount}
\usepackage{abc}
\usepackage{tocloft}
\usepackage[utf8]{inputenc}
\usepackage[top=48pt,headheight=18pt,headsep=12pt,bottom=48pt,inner=48pt,outer=48pt]{geometry}
\makeindex

\begin{document}
%\setuplayouts
\frontmatter

\pagestyle{fancy}

\fancyhf{}

\fancyhead[LE,RO]{\thepage}
\fancyhead[RE,LO]{DRAFT \today -- {\bf DO NOT DISTRIBUTE}}
%\pagebreak
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}
\setlength{\parindent}{0pt}
\setlength{\parskip}{11pt plus 2pt minus 2pt}

%\begin{figure}
%
%\currentpage
%
%\drawparametersfalse
%
%\drawpage
%
%\caption{Page layout for this document} \label{fig:ptrs}
%
%\end{figure}
%\pagebreak
%

\section*{Note from the Editor}

The goal of this book is to include as many tunes as possible for extant
pre-1651 dances. It also includes music for dances choreographed by SCA
memebers in a variety of styles as well as some tunes for later English Country
dances and modern folk dances that are danced in the SCA.

We have provided this fakebook both as a printable PDF and as ABC files,
obtainable from \url{https://github.com/orgs/st-cecilia-press/rendance\_abc}.
ABC is a simple format for music readable both by computers and by humans that
is commonly used for folk music and dance tunes. There is also a lot of
software to display, play, search, and transpose ABC files. We particularly
recommend EasyABC on Mac, Linux and Windows:
\url{http://www.nilsliberg.se/ksp/easyabc/}. There are many other software
packages including for Android and iOS devices. A full list can be found at
\url{http://abcnotation.com/software}.

We have aimed to match the key of the tune as well as the marked chords to a
tune's most recent appearance in the Pennsic Pile. If it has not recently
appeared in the Pennsic Pile, we have typically kept the original key in the
source material and added an attempt at a reasonable harmonization.

Our belief is that all these tunes are freely usable within the SCA. Most are
many hundreds of years old. However, please make your own determination before
using any of this material in a non-SCA/non-educational setting. Sources of
harmonizations are noted in the ABC files mentioned above.

Some of these tunes have many more chords marked than most chord players would
reasonably want to play. These extra chord markings may still be useful for
players improvising a bass line or for adding passing notes when playing a
chordal accompaniment.

Thanks to Emma Badowski for organizing this material in a previous edition of
this fakebook and especially for harmonizing many of the tunes from Playford
1651 that have not previously appeared in the Pennsic Pile.

Please note that this book is {\em not} an official publication of the Society
for Creative Anachronism.

Aaron Elkiss

\clearpage
\begin{multicols}{2}

\renewcommand\cftchapafterpnum{\vskip\baselineskip}
\setlength{\cftsubsecindent}{0pt}
\setlength{\cftsubsecnumwidth}{0pt}
\tableofcontents
\end{multicols}

\clearpage
\mainmatter

\chapter{Basse Danse}

Basse danse (or bassadanza in Italian) was popular across Europe in the 15th
and early 16th centuries. One of the most important sources for basse danse is
Ms 9085 in the Bibliotheque Royale, Brussels (c. 1445) This manuscript gives
only a slow-moving tenor, or cantus firmus, as the melody for most of the
dances. Musicians normally would have improvised multipart polyphony above the
tenor line. One simple way to improve a melody from the tenor line is to play
it in the style of the basse dance section from ``Rostiboli Gioioso''.  Most of
the basse danses are notated here in 6/4 time, and an appropriate tempo would
be approximately dotted half note = 40-45.

\clearpage

\input{basse_dance.tex}

\chapter{15th Century Italian Dances}

The primary sources for 15th Century Italian dance are manuscripts from the
mid- to late 15th century containing dances by (among others) the dancing
masters Domenico da Piacenza (c. 1400-1470) and his student Guglielmo Ebreo (c.
1420-1848) (also known as Giovanni Ambrosio after his conversion from Judaism
to Catholocism).

15th century Italian dance is somewhat unusual in that dances often change
between ``tempi'', which are marked in each dance. The various tempi are
transcribed as:

\begin{itemize}

\item Bassadanza: 6/4
\item Quadernaria: 4/4
\item Saltarello: 6/8 or occassionally 3/4
\item Piva: 2/4 or 6/8

\end{itemize}

As a rough guide for tempo, keeping a constant tempo of approximately quarter
note = 120 (or dotted quarter = 120 for 6/8 piva sections) regardless of the
various tempi should work for many of the dances.

(See {\em Joy and Jealousy} by Vivian Stephens and Monica Cellio for additional
information; it is available online at
\url{http://sca.uwaterloo.ca/~praetzel/Joy-J-book/}). 


\clearpage
\input{15italian.tex}

\chapter{Dances from the Gresley Manuscript}

The Gresley manuscript dates to the late 15th or early 16th century and was
re-discovered in Derbyshire, England. It contains choreography for 26 dances
and music for 13, with 8 of those having both music and the dance steps. We
have re-used other music from the manuscript for some of the dances missing
music and have included newly-composed music by Master Martin Bildner for the remainder, provided
under a Creative Commons Attribution-NonCommercial-ShareAlike license (see 
\url{https://creativecommons.org/licenses/by-nc-sa/3.0/}).
Reconstructions very, so always check the music with the dance master!

The dances are primarily transcribed in a lively 6/8 time; a tempo of dotted
quarter = 115-120 should work well.

\clearpage
\input{gresley.tex}

\chapter{Dances from the Inns of Court}

The dances in this section are from the Inns of Court: professional
associations for English barristers dating to the 15th century. There are
several known manuscripts dating from the mid-16th to mid-17th century
informally describing these dances, eight of which are believed to have been
performed in a fixed order at the beginning of revels at the Inns of Court. We
have preserved that order (for Quadran Pavane through Black Alman) to
facilitate dancing the entire suite, also known as ``The Old Measures''.

Tempos vary wildly, so check with the dancing master for their preference.
Reconstructions vary as well, so check for the desired roadmaps especially for
the more unusual ones such as Turkelone and Tinternell. We have included
suggested numbers of repeats when playing all 8 Old Measures as a suite, but
you may want to play more times through the dance if playing just one of the
dances.

\clearpage
\input{inns.tex}

\chapter{16th Century Italian Dances}

The major sources for 16th century Italian dances are the published books of
Fabritio Caroso (c. 1526-1605) and Cesare Negri (c. 1535-1605).

Many of the dances included in this collection are {\em cascarda}, a bouncy,
triple time kind of dance unique to Caroso. We have used a 3/4 time signature
for these but the dances should really be felt in 1, with a tempo of
approximately dotted half = 110-120.

The other dances (mostly {\em balletti}) in common time such as Bizzarria and
Lo Spagnoletto should work well with a tempo of half note = 100-110. Some of
these dances shift to 3/4 time partway through; let dotted half note in the 3/4
section = half note in the common time section. 

A few exceptions: Passo e Mezzo is written with doubled note values in cut
time, so use a tempo of whole note = 100-110. There are also a few dances we
have transcribed in 3/4 that are not cascarda such as Contrapasso and
Villanella. For Contrapasso, use a tempo of dotted half = 50-55. For
Villanella, always check with the dance master: it is sometimes danced (at the
same speed) to the music played slowly for 3 repeats and sometimes to the music
played twice as fast for 6 repeats. 

\clearpage
\input{16italian.tex}

\chapter{Dances from Arbeau's {\em Orchésographie}}

Published in 1589 in Langres, France, Orchésographie includes music and
instructions for many different kinds of dances. Numerically speaking, the bulk
of the dances in Arbeau are {\em bransles}. Most of the bransles are in duple
time and should be played at about half note = 115.  The triple time bransles
are Bransle Gay and Bransle de Poictou; for these, a tempo of dotted half =
60-65 should work.

Arbeau also includes instructions for the pavane and galliard, music for which
also appears in the Improvised Dances section.

\clearpage
\input{arbeau.tex}

\chapter{Improvised Dances}

Improvised dances such as the pavane and galliard were very popular in the 16th
century all over Europe. Music and instructions for these dances appear in
numerous sources. Reductions are provided from such sources as Tylman Susato's
{\em Danserye} of 1551 and Praetorius' {\em Terpsichore} of 1612. We have also
included the tunes used in the SCA for some early Italian improvised dances,
the Piva and the Saltarello.

The Canarie is transcribed in 6/4. For the Canarie, use a tempo of
approximately dotted half = 70.

Galliards can be transcribed in either 3/2 or 6/4. We have chosen to use 3/2
for clarity for some of the more rhythmically complex settings while halving
the original note values and using 6/4 for the more straightforward ones. The
tempo for galliards (for the 6/4 settings) can be anywhere from dotted half =
45 - 60, depending on the whims of the dancing master. For transcriptions in
3/2 use dotted whole = 45 - 60 instead (two measures of a 3/2 galliard equating
to one measure of 6/4 galliard). The Volta is really just a variation on the
galliard and can be played as such.

The pavanes are transcribed in cut time, and again, the tempo can range from
half note = 45 to 60.

Preferences vary, so always check with the dancing master for desired tempo.
Additionally, modern choreographies have been created for some of these tunes,
so be sure to confirm the roadmap with the dancing master if these are being
danced.

\clearpage
\input{improvised.tex}

\chapter{The English Dancing Master, 1651}

This section includes all 105 dances in the first edition of John Playford's
{\em The English Dancing Master} of 1651.  The dances are generally transcribed
in either cut time or in 6/4. For cut time use a tempo of approximately half
note = 115 or for 6/4, dotted half = 115.  Some dances such as Chestnut are
often danced slower, so be sure to check with the dancing master just in case.

\clearpage
\input{playford.tex}

\chapter{Other English Country Dances}

This section includes a variety of other English Country dances including
reconstructed music for dances from pre-Playford manuscripts as well as
selected dances from later editions of Playford or later English Country dance
authors that have been danced in the SCA.

\clearpage
\input{playford_later.tex}

\chapter{Modern Folk Dances and SCA Choreography}

This section includes folk dances from later traditions that are popular within
the SCA as well as music for SCA choreographies in a variety of styles.

\clearpage
\input{other.tex}

\clearpage

\printindex

\end{document}
